---
title: "Homework 2"
output: github_document
date: "2026-10-03"
---

```{r setup}
library(tidyverse)

```

# Problem 1

## Importing and cleaning the data
```{r q1_import_clean}
nyc_subways = 
  read_csv(
    file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"
  ) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

nyc_subways
```

The NYC subway dataset contains information on each entrance and exit for each subway station in NYC. After cleaning, the dataset has `r nrow(nyc_subways)` rows and `r ncol(nyc_subways)` columns. The variables retained are: `r paste(names(nyc_subways), collapse = ", ")`. To clean the data, I cleaned the variable names using `janitor::clean_names()` to make them easier to work with, selected the variables of interest (line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance), and converted the `entry` variable from character (YES / NO) to logical (TRUE / FALSE). These data are not tidy: route information is spread across the eleven columns `route1` through `route11` (with `route8` through `route11` as numeric variables and `route1` through `route7` as character variables) so the route is stored in column names rather than as a single variable with one value per row.

## To answer the questions:
```{r q1_questions}
n_stations = 
  nyc_subways |> 
  distinct(line, station_name) |> 
  nrow()
 
n_ada = 
  nyc_subways |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()
 
prop_no_vending = 
  nyc_subways |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

There are `r n_stations` distinct stations, identified by both line and station name.
`r n_ada` of these stations are ADA compliant.
The proportion of station entrances / exits without vending that allow entrance is `r round(prop_no_vending, 3)`.

## Reformatting
```{r q1_reformatting}
# First, have to make sure that all of the route variables are character.
nyc_subways = 
  nyc_subways |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  )

# using pivot_longer() to make one route number column and one route name column
nyc_subways_longer = 
  nyc_subways |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

# To answer the questions about the A train:
n_a_stations = 
  nyc_subways_longer |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada = 
  nyc_subways_longer |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
```

`r n_a_stations` distinct stations serve the A train. Of these, `r n_a_ada` are ADA compliant.

# Problem 2
