---
title: "Homework 2"
output: github_document
date: "2026-10-03"
---

```{r setup}
library(tidyverse)

```

# Problem 1

## Importing and cleaning the data
```{r q1_import_clean}
nyc_subways = 
  read_csv(
    file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"
  ) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

nyc_subways
```

The NYC subway dataset contains information on each entrance and exit for each subway station in NYC. After cleaning, the dataset has `r nrow(nyc_subways)` rows and `r ncol(nyc_subways)` columns. The variables retained are: `r paste(names(nyc_subways), collapse = ", ")`. To clean the data, I cleaned the variable names using `janitor::clean_names()` to make them easier to work with, selected the variables of interest (line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance), and converted the `entry` variable from character (YES / NO) to logical (TRUE / FALSE). These data are not tidy: route information is spread across the eleven columns `route1` through `route11` (with `route8` through `route11` as numeric variables and `route1` through `route7` as character variables) so the route is stored in column names rather than as a single variable with one value per row.

## To answer the questions:
```{r q1_questions}
n_stations = 
  nyc_subways |> 
  distinct(line, station_name) |> 
  nrow()
 
n_ada = 
  nyc_subways |> 
  filter(ada) |> 
  distinct(line, station_name) |> 
  nrow()
 
prop_no_vending = 
  nyc_subways |> 
  filter(vending == "NO") |> 
  pull(entry) |> 
  mean()
```

There are `r n_stations` distinct stations, identified by both line and station name.
`r n_ada` of these stations are ADA compliant.
The proportion of station entrances / exits without vending that allow entrance is `r round(prop_no_vending, 3)`.

## Reformatting
```{r q1_reformatting}
# First, have to make sure that all of the route variables are character.
nyc_subways = 
  nyc_subways |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  )

# using pivot_longer() to make one route number column and one route name column
nyc_subways_longer = 
  nyc_subways |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

# To answer the questions about the A train:
n_a_stations = 
  nyc_subways_longer |> 
  filter(route_name == "A") |> 
  distinct(line, station_name) |> 
  nrow()

n_a_ada = 
  nyc_subways_longer |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name) |> 
  nrow()
```

`r n_a_stations` distinct stations serve the A train. Of these, `r n_a_ada` are ADA compliant.

# Problem 2
## Importing and cleaning datasets
```{r q2_trash_import}
# separate sheets for the trash wheels
mr_trash_df =
  readxl::read_excel("data/202509 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N709") |> #using the range instead of skip because of some notes coming in through the excel
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    sports_balls = as.integer(round(sports_balls, 1)),
    wheel = "Mr. Trash Wheel",
    month = str_to_lower(month)
  )

prof_trash_df =
  readxl::read_excel("data/202509 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    wheel = "Professor Trash Wheel",
    month = str_to_lower(month)
  )

gwynn_df =
  readxl::read_excel("data/202509 Trash Wheel Collection Data.xlsx", sheet = "Gwynns Falls Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    wheel = "Gwynnda",
    month = str_to_lower(month)
  )

# combining into one trash data frame
trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynn_df) |> 
  relocate(wheel)
```

The `trash_df` dataframe combines the data from three trash wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. After cleaning and combining the three datasets, the final dataset contains `r nrow(trash_df)` observations with `r ncol(trash_df)` variables. The variables retained are: `r paste(names(trash_df), collapse = ", ")`. Key variables include `wheel`, which identifies the trash wheel, `weight_tons` and `volume_cubic_yards` for each dumpster, and counts of items collected such as `plastic_bottles`, `cigarette_butts`, and `wrappers.` Not every item is recorded for every wheel: `sports_balls` is only recorded for Mr. Trash Wheel and `glass_bottles` is not recorded for Gwynnda, so these are missing for the other wheels.

The total weight collected by Professor Trash Wheel is `r trash_df |> filter(wheel == "Professor Trash Wheel") |> pull(weight_tons) |> sum(na.rm = TRUE)`. Gwynnda collected a total of `r trash_df |> filter(wheel == "Gwynnda", month == "june", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE) |> format(big.mark = ",")` cigarette butts in June of 2022.
