---
title: "Homework 2"
output: github_document
date: "2026-10-03"
---

```{r setup}
library(tidyverse)

```

# Problem 1

## Importing and cleaning the data
```{r q1_import_clean}
nyc_subways = 
  read_csv(
    file = "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv"
  ) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, route1:route11, entry, vending, entrance_type, ada) |> 
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

nyc_subways
```

The NYC subway dataset contains information on each entrance and exit for each subway station in NYC. After cleaning, the dataset has `r nrow(nyc_subways)` rows and `r ncol(nyc_subways)` columns. The variables retained are: `r paste(names(nyc_subways), collapse = ", ")`. To clean the data, I cleaned the variable names using `janitor::clean_names()` to make them easier to work with, selected the variables of interest (line, station name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance), and converted the `entry` variable from character (YES / NO) to logical (TRUE / FALSE). These data are not tidy: route information is spread across the eleven columns `route1` through `route11` (with `route8` through `route11` as numeric variables and `route1` through `route7` as character variables) so the route is stored in column names rather than as a single variable with one value per row.

## To answer the questions:
```{r q1_questions}
stations = 
  nyc_subways |> 
  distinct(line, station_name)
 
ada = 
  nyc_subways |> 
  filter(ada) |> 
  distinct(line, station_name)
 
no_vending = 
  nyc_subways |> 
  filter(vending == "NO") |> 
  pull(entry)
```

There are `r nrow(stations)` distinct stations, identified by both line and station name.
`r nrow(ada)` of these stations are ADA compliant.
The proportion of station entrances / exits without vending that allow entrance is `r round(mean(no_vending), 3)`.

## Reformatting
```{r q1_reformatting}
# First, have to make sure that all of the route variables are character.
nyc_subways = 
  nyc_subways |> 
  mutate(
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
  )

# using pivot_longer() to make one route number column and one route name column
nyc_subways_longer = 
  nyc_subways |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )

# To answer the questions about the A train:
a_stations = 
  nyc_subways_longer |> 
  filter(route_name == "A") |> 
  distinct(line, station_name)

a_ada = 
  nyc_subways_longer |> 
  filter(route_name == "A", ada) |> 
  distinct(line, station_name)
```

`r nrow(a_stations)` distinct stations serve the A train. Of these, `r nrow(a_ada)` are ADA compliant.

# Problem 2
## Importing and cleaning datasets
```{r q2_trash_import}
# separate sheets for the trash wheels
mr_trash_df =
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N742") |> #using the range instead of skip because of some notes coming in through the excel
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel",
    month = str_to_lower(month)
  )

prof_trash_df =
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", skip = 1) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles), #Dumpster 135 had a plastic_bottles value of "dive". as.numeric() converts this to NA
    wheel = "Professor Trash Wheel",
    month = str_to_lower(month)
  )

gwynn_df =
  readxl::read_excel("data/202610 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda the Good Wheel of the W", skip = 1) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.numeric(year),
    wheel = "Gwynnda",
    month = str_to_lower(month)
  )

# combining into one trash data frame
trash_df = 
  bind_rows(mr_trash_df, prof_trash_df, gwynn_df) |> 
  relocate(wheel)
```

The `trash_df` dataframe combines the data from three trash wheels: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. After cleaning and combining the three datasets, the final dataset contains `r nrow(trash_df)` observations with `r ncol(trash_df)` variables. The variables retained are: `r paste(names(trash_df), collapse = ", ")`. Key variables include `wheel`, which identifies the trash wheel, `weight_tons` and `volume_cubic_yards` for each dumpster, and counts of items collected such as `plastic_bottles`, `cigarette_butts`, and `wrappers.` Not every item is recorded for every wheel: `sports_balls` is only recorded for Mr. Trash Wheel and `glass_bottles` is not recorded for Gwynnda, so these are missing for the other wheels.

The total weight collected by Professor Trash Wheel is `r trash_df |> filter(wheel == "Professor Trash Wheel") |> pull(weight_tons) |> sum(na.rm = TRUE)`. Gwynnda collected a total of `r trash_df |> filter(wheel == "Gwynnda", month == "june", year == 2022) |> pull(cigarette_butts) |> sum(na.rm = TRUE) |> format(big.mark = ",")` cigarette butts in June of 2022.

# Problem 3

## Baseline Data File
```{r q3_baseline_file}
# full file (includes people who are excluded from the study)
baseline_df_raw = 
  read_csv("data/data_mci/MCI_baseline.csv", skip = 1, na = c("", "NA", ".")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex, 1 ~ "Male", 
      0 ~ "Female"
    ),
    apoe4 = case_match(
      apoe4, 
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  )

# excluding those with MCI at baseline
baseline_df = 
  baseline_df_raw |> 
  filter(is.na(age_at_onset) | age_at_onset > current_age)
```

```{r q3_baseline_descriptives}
# These codes are for the descriptives that are not as easy to show in just inline code

# for number who developed MCI
mci_bl =
  baseline_df |> filter(!is.na(age_at_onset))

#for the proportion of women who are APOE4 carriers
women_apoe4 = 
  baseline_df |> 
  filter(sex == "Female") |> 
  mutate(
    carrier_logic = case_match(
      apoe4,
      "carrier" ~ TRUE,
      "non-carrier" ~ FALSE
    )
  )
```

The `baseline_df` data frame contains contains baseline information on `r nrow(baseline_df)` patients without MCI at baseline. To create this dataset, I imported the raw file, defined the NA values, skipped the first line of information about the dataset, cleaned the names, recoded sex and APOE4 so they were not numeric using the key that I dropped on import, and filtered out the patients that had MCI at baseline. A total of `r nrow(baseline_df_raw)` patients were recruited, of which `r nrow(baseline_df)` did not have MCI at onset (met inclusion criteria and were thus eligible). Of these patients, `r nrow(mci_bl)` developed MCI. 

The average age at baseline of these patients is `r round(mean(pull(baseline_df, current_age)), 2)` and the proportion of women who are APOE4 carriers in the study is `r mean(pull(women_apoe4, carrier_logic))`. 

## Longitudinally-Observed Biomarkers Dataset
```{r}
amyloid_df = 
  read_csv("data/data_mci/mci_amyloid.csv", skip = 1, na = c("", "NA", ".", "Na")) |> 
  janitor::clean_names() |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio"
  ) |> 
  mutate(
    time = replace(time, time == "baseline", 0), 
    time = as.numeric(time),
    amyloid_ratio = as.numeric(amyloid_ratio)
  ) |> 
  rename(id=study_id)
```

For the dataset of longitudinally observed biomarker values, I imported the data file, defined NAs (specifically, defined "Na" as an NA value), cleaned the names,  pivoted the time columns, and converted the time and amyloid ratio columns to numeric values. This dataset has `r nrow(amyloid_df)` rows and three columns (study id, where biomarker Amyloid 42/40 ratio was measured. The average amyloid ratio is `r round(mean(pull(amyloid_df, amyloid_ratio), na.rm = TRUE), 3)`. 


## Comparing the values in the two datasets
```{r q3_comparison}
# data frames for those who only show up in one of the two datasets
baseline_only_df = 
  baseline_df |> 
  filter(!id %in% pull(amyloid_df, id))

amyloid_only_df = 
  amyloid_df |> 
  filter(!id %in% pull(baseline_df, id)) |> 
  distinct(id)

# combined dataset of those in both
mci_df = 
  baseline_df |> 
  filter(id %in% pull(amyloid_df, id)) |> 
  left_join(amyloid_df, by = "id")

write_csv(mci_df, "data/mci_combined.csv")
```

Comparing the two datasets, `r nrow(baseline_only_df)` participants appear only in the baseline data (so they have no amyloid measurements), and`r nrow(amyloid_only_df)` appear only in the amyloid data. The combined dataset retains the `r mci_df |> distinct(id) |> nrow()` participants who appear in both, with `r nrow(mci_df)` rows (one per participant per visit) and `r ncol(mci_df)` variables: `r paste(names(mci_df), collapse = ", ")`. `r sum(is.na(pull(mci_df, amyloid_ratio)))` rows have a missing amyloid ratio.